\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-30 paper/main.tex

\todo[inline]{Likely discussion:}
The aim of \name is to predict candidate drugs that target a given
protein; the challenge is to develop a training and evaluation scheme
that does not simply overfit to the inherent biases in training and
testing data.
In general, when performing cross-validation for DTI prediction, the
options are to split over 
\begin{enumerate}
	\item split over drugs,
	\item split over drug--target pairs, or
	\item split over proteins
\end{enumerate}
where the first and third option concern splitting drugs and proteins, respectively, into train, validation and test sets, and arranging the corresponding drug-target interactions. They that at least parts of the interactions are not seen during training and evaluate either how well targets are predicted for unseen drugs or unseen proteins. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. \\

\todo[inline]{Likely discussion:}
The most common scheme for DTI prediction is the split over drug-target pairs \citep{Survey2018}, where likely all drugs and targets of the validation and testing phase have already occurred in the training phase, as part of other drug-target pairs. The second most prevalent arrangement is the split over drugs, while only close to none is aiming on a protein split.  However, the first and second splitting scheme are exposed to the first dataset bias and are hence more likely vulnerable to transductive inference by just predicting recently seen structures, rather than implementing inductive inference and generalizing over the drug representations. Second, these two strategies are more susceptible to the second and third bias, as only in these cases the model may overfit on the number of existing interactions for a single protein, while in the third scheme the number of interactions of the test proteins is not known.

@@

Our aim is to develop and evaluate a computational model to predict
drug--target interactions. Specifically, given a protein, we are
interested in identifying the drugs that may target the protein; this
scenario is motivated by the need to identify drugs that can target a

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two 
types of features: structural information for drugs and proteins that
can be used to determine if they may physically interact, and
information about drug effects that may localize on an interaction
network.

This scenario is motivated by the need to identify drugs that can target a
specific protein that may be associated with a disease or abnormal
phenotypes.
Our aim is further to combine two different types of features:
structural information based on protein structure and molecular drug composition, and their respective observable characteristics.

@@

To obtain a prediction for a given drug-protein pair\todo{should this
  be about pairs, or rather something like ``given a protein, we aim
  to find a drug...''? I.e., find a consistent way to separate
  yourself language-wise from a naive prediction of pairs.}, we first
build DL2vec representations\todo{In results: talk about what they are
supposed to DO (``embeddings/representations of the phenotypes and
functions...''), not how you implement them (you could use something
else, for example).} for both drugs and proteins over PhenomeNET, while also preparing structural, molecular representations. Hence, we pretrain the siamese networks for both molecular and ontology features, yielding the deep neural network transformers for both methods and both drugs and proteins, individually. These embeddings are now added to the graph neural network as node features, where we start an end-to-end learning and training process. Notably, like in the pretraining phase, we rely on siamese neural networks, enforcing similar representations for fitting drug-protein pairs. The overall architecture can be seen in Figure~\ref{fig:Fullmodel}.

\subsection{High level description}

For a given protein, we want to find all suitable interacting drugs, \textbf{among a set of known drugs}.
To obtain a prediction for a given drug-protein pair, we first
build representations embodying both phenotypes and functions for both drugs and proteins, while also preparing structural, molecular representations, e.g., utilizing DL2vec over PhenomeNET and the prebuilt features from SMILES transformer and DeepGOPlus. Note that this approach can be easily generalized, profiting from other protein function and phenotype representation methods. \\ 
Hence, we pretrain the siamese networks for both molecular and ontology features, yielding the deep neural network transformers (DNNT) for both methods, and both drugs and proteins, individually. These embeddings are now added to the graph neural network as node features, where we start an end-to-end learning and training process. Notably, like in the pretraining phase, we rely on siamese neural networks, enforcing similar representations for fitting drug-protein pairs. The overall architecture is depicted in Figure~\ref{fig:Fullmodel}.\\

>>>>

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank known drugs that likely target this protein. \name combines two 
types of features: structural information for drugs and proteins that
can be used to determine if they may physically interact, and
information about drug effects that may localize on an interaction
network.
As structural features, \name uses structural representations of drugs
from the SMILES transformer \cite{SmilesTransformer} and representations of protein
amino acid sequences from DeepGOPlus \cite{DeepGoPlus}.
\name learns representations of drug effects and protein functions
using the ontology-based machine learning method DL2Vec \cite{DL2vec2020}.

We use two Siamese neural networks to combine the molecular
and phenotype representations of drugs and proteins,
% and
% ontology features, yielding the deep neural network transformers
% (DNNT) for both methods, and both drugs and proteins,
% individually. 
and add these to a protein interaction network as node features.
% neural network in which nodes representing interactions
% between proteins as node features, 
We then train this model in an end-to-end manner. % Notably, like in the
% pretraining phase, we rely on siamese neural networks, enforcing
% similar representations for fitting drug-protein pairs. 
The overall architecture is depicted in Figure~\ref{fig:Fullmodel}.

% For a given protein, we want to find all suitable interacting drugs, \textbf{among a set of known drugs}.
%To obtain a prediction for a given drug-protein pair,

Note that this approach can be easily generalized, profiting from
other protein function and phenotype representation methods.

\todo[inline]{Likely discussion:}
The aim of \name is to predict candidate drugs that target a given
protein; the challenge is to develop a training and evaluation scheme
that does not simply overfit to the inherent biases in training and
testing data.
In general, when performing cross-validation for DTI prediction, the
options are to split over 
\begin{enumerate}
	\item split over drugs,
	\item split over drug--target pairs, or
	\item split over proteins
\end{enumerate}
where the first and third option concern splitting drugs and proteins, respectively, into train, validation and test sets, and arranging the corresponding drug-target interactions. They ensure that at least parts of the interactions are not seen during training and evaluate either how well targets are predicted for unseen drugs or unseen proteins. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. \\

\todo[inline]{Likely discussion:}
The most common scheme for DTI prediction is the split over drug-target pairs \citep{Survey2018}, where likely all drugs and targets of the validation and testing phase have already occurred in the training phase, as part of other drug-target pairs. The second most prevalent arrangement is the split over drugs, while only close to none is aiming on a protein split.  However, the first and second splitting scheme are exposed to the first dataset bias and are hence more likely vulnerable to transductive inference by just predicting recently seen structures, rather than implementing inductive inference and generalizing over the drug representations. Second, these two strategies are more susceptible to the second bias, as only in these cases the model may overfit on the number of existing interactions for a single protein, while in the third scheme the number of interactions of the test proteins is entirely unknown during training process.

@@

@@
